Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies

### 2019-03-14 man/ms.Rmd
update bib and ms

Although our sampled sited showed similar tree competence levels (Table 1), the review of historical documents showed that la historia de manejo humano ha sido diferente para cada sitio. Different values of tree-growth were observed between populations and for different elevations within southern location (Figure 4). Low-elevation sites showed lower BAI values than higher-elevation site. A reason that could explain these results is the different degree of anthropic impact related to elevation. Los bosques situados a mas elevación presentan menor presión antrópica que los localizados a menor altitud, por el simple hecho de que estos últimos están mas accesibles. During the last decades, the rural abandonment has lead an decrease of anthropic pressures on the high-elevations zones of Mediterranean mountains (INLCUIR CITA).

comentario --> Aqui pretendo incluir la idea que hablamos en su momento. Quizá están creciendo donde les están dejando crecer (menor impacto antrópico en las zonas mas altas, al menos en los últimos años -abandono de zonas rurales?-)
He ralizado una revisión de trabajos que utilicen dendro de Q. pyrenaica para ver si existe una relación entre la edad de las cronos con la elevación.

* Son bosques relativamente jóvenes 
comentaio --> 
Aquí me falta desarrollar este apartado algo más incluyendo cosas de los documentos históricos (la minería, etc, etc) 
Los tres sitios mostraron una reducción en los valores de BAI durante las sequías de 1995 y 2005, y menos evidente para el año 2012 (Figure 5). La población del norte (SJ) mostró una gran liberación en torno a 1945 fruto quizá de cortas asociadas a la actividad minera de la zona. Asimismo, esta población mostró un marcado descenso en BAI durante el periodo 1986 - 1999.

>>>>

The Mediterranean mountains, even at high elevations, have suffered strongly transformation of the landscapes driven by human activities throughout history [@Regato2008]. Multiple evidence from palaeoecological studies indicated an intense human impact on vegetation of Sierra Nevada since 3000 cal year BP [@Anderson2011; @JimenezMoreno2013; @GarciaAlix2017, @MesaFernandez2018]. From that moment onwards, increases on fire occurrence and both grazing and mining activities, were recorded for this mountain region, with an sharp intensification of the human activities in the last 150 years [@GarciaAlix2017; @MesaFernandez2018]. Mining, forest clearings, fuelwood and charcoal exploitations, pastoralism and also wars, have strongly impacted on the forest resources in Sierra Nevada. As a consequence of all those activities a loss of about 90% of broadleaf *Quercus* species cover from medium and low elevation occurred in this mountain region [@JimenezOlivencia1991].

Our study sites have been subjected to an intense exploitation of forest resources. Although the estimated age for our sampled trees was not more than 180 years (Tables 1, 2), several documents reported the presence of oaks before. For instance, the inventories of trees made by the Spanish Navy during the second half of 18th century, recorded the quantity of trees, dividing them into three categories: *new*, *growing* and *old* [@Wing2015]. For Cáñar site more than two millions of trees were reported, most of them *news*, and no *old* trees were counted, suggesting recent wood fellings. Less quantity were reported for San Juan location (circa 700 000 trees), which 220 *old* trees and 56 700 *growing* [@GomezCruz1992]. These numbers are a reflect of the intense exploitation of the forest resources occurring up to the end of nineteenth century [@ValbuenaCarabana2010; @Calatrava2019] and also concur with the regeneration peaks observed for several Iberian oak woodlands [e.g. @GeaIzquierdo2014; GeaIzquierdo2015; @Dorado2017].

We found similar tree competence levels in our sampled sites (Table 1), but differences in tree size and age suggest different management origin (*i.e.* land-use legacy). Our study sites had different land-use trajectories driven by the differential pattern of the natural resource use [@JimenezOlivencia2015; @MartinCivantos2014].

In Northern sites (*e.g.* San Juan), there is a sequentially distribution of the land uses along the elevational gradient [CITA CATALOGO]. Grasslands and shrublands for cattle farming are located at high elevations. Then forests formation with some croplands; and then, at lowlands, irrigated terraces with tree crops. Another activities could conditioning the natural resources uses. For instance, San Juan oak woodland is located in an area with a high concentration of mines and quarries that have been exploited intermittently throughout history (citas), with two periods of intense activity, the second half of the 19th century after the publication of detailed mineralogical reports by [@MaestreXXXX]; and during the first decades of the twentieth century until 1960, which is the last year in which there is evidence of the existence of mining activity [cita]. The mining activity in San Juan site have strongly conditioned the forest structure.

On the other hand, woodland areas of the southern slopes (*e.g.* Cáñar) are mixed with a greater percentage of croplands, even reached high elevation (mainly barley, rye and potatoes)[@Calatrava2019]. Forest resources (firewood, charcoal, acorns) were used in a continuously way, as some historical sources have pointed out [@ValbuenaCarabana2014, @JimenezSerrano2004]. In fact, in Cáñar there were an arraigated tradition of charcoal extraction since fifteenth century [@JimenezSerrano2004] to late twenty century [@MesaTorres2009; @Bonet2014_conama].

The different patterns of resource uses can help to understand the tree-growth changes recorded in our chronologies (Figure 5). We observed a tree-growth release event around 1940 in the SJ oak woodland (Figures 5, S4) which concurs with one period of mining activity for this area. Several documentary sources indicate an apogee of the mining activity during 1925 to 1957 period, supported by both the improvement of the paths to transport the mineral and by the creation of new railway connections (CITA). During this period there were an increase on the use of timbers for the tunnels of the mines (CITE) and several furnaces that required great quantities of fuelwood to melt the mineral were active in this area (cite). This heavily exploitation of the forest resources could affect to a major part of this oak woodland, since percentage of trees affected by GC > 50 % reaches values above 50 % (Figure S4). Our results also coincide with a concurrent but less intense tree-growth release event reported for a closed oak woodland [@GeaIzquierdo2014]. In addition, paleoecological studies carried out in nearby alpine bogs have recorded increases in the heavy metals concentration since the end of 18th century until mid-20th, which coincides with the maximum activity of the mining in this area [@GarciaAlix2017].

The release event observed for SJ site during 1995-2000 was lower than the occurred at 1940, but affecting more trees (Figures 5, S4). The revision of forests practices carried out in this area in the last 30 years did not reveal any clearing or cutting event for those years. Two non-mutually exclusive ways could explain the tree-growth release observed. Firstly it could be related with a natural drought-induced mortality event after 1995, as was reported for Mediterranean tree species [*e.g.* @Penuelas2000; @Lloret2004; @Gentilesca2017]. On the other hand, the strong positive correlations of SPEI with tree-growth obtained for this site (Figure 8) indicates a high sensitivity to water availability of this slightly more xeric site (Table 1) [@GeaIzquierdo2014]. Previous works reported a sharp decline (or non-production) in latewood production during extreme drought events [@Corcuera2006]. Since latewood are less vulnerable to embolism than earlywood vessels [@Corcuera2006], the reduction or even the non-production of former could negatively affect to tree-growth and also enhanced the mortality, particularly for drier sites [@Corcuera2006; @GeaIzquierdo2014].
